Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-28 doc/thesis/results.tex
Draft???

The \gls{alpha} parameter was the most accurately estimated, with point
estimates having an average deviation of 
    0.14
from the true value, on a grid from 0 to 2. The error was negatively correlated
with the true value of \gls{alpha} 
    (Spearman's $\rho$ = \ensuremath{-0.26},
     $p = 1\!\times\!10^{-5}$),
although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
The accuracy was highest for the test value \gls{alpha} = 1.25
    (mean error 0.02)
which exhibited markedly different behaviour than the other values in terms of
the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
particular, there was a very pronounced peak in scores around the true value,
in contrast to most other values where the scores were flat around the true
value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
average absolute error of the point estimates for \gls{I} was 
    310,
on a grid of 500 to 5000, and the errors were not significantly correlated with
the true value of \gls{I}. Kernel score distributions for all test values
exhibited a similar rounded shape (\cref{fig:gridI}).

@@

We used \software{netabc} to estimate the parameters of the \gls{BA} model on
simulated trees where the true parameter values were known. Point estimates for
each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
2. The results for the other values of \gls{m} were similar
(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates of
\gls{alpha} across all simulations was
    0.08 
    [0.05-0.17].
The accuracy of the estimates was not significantly different between values of
$m$ or $I$ (both one-way ANOVA,
    $p$ = 0.05
and 
    0.07),
although the errors when the true value of \gls{alpha} was zero were
significantly greater than the other values 
    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$).
The error in the estimated value of $I$ was
    395 
    [207-683].
Errors were significantly higher for $\alpha \geq 1$
    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$)
and for $I$ = 2000
    ($p = {<}10^{-5}$),
but not for any values of $m$ (one-way ANOVA). The $m$ parameter was estimated
correctly in only
    27 \%
of simulations. The true values of $m$ and $I$ did not significantly affect the
error (one-way ANOVA), but the accuracy was significantly lower for integral
than non-integral values of \gls{alpha}
    (Wilcoxon rank-sum test, $p = 0.3$).
Finally, the total number of nodes \gls{N} was consistently over-estimated by
about a factor of two
    (error 5987 
    [2060 - 
     7999]).
No other parameters influenced the accuracy of the $N$ estimates (one-way
ANOVA).

@@

\begin{figure}
  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
  \caption[Approximate marginal posterior distributions of BA model parameters
      obtained by applying \textit{netabc} to a simulated transmission tree
      with values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
    {
        Approximate marginal posterior distributions of BA model parameters
        obtained by applying \textit{netabc} to a simulated transmission tree
        with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
        highest posterior density intervals. $x$-axes indicate regions of
        nonzero prior density.
    }
  \label{fig:abcex}
\end{figure}

\begin{table}
    \centering
    \input{\tablepath/abc-hpd}
    \caption{Average widths of 95\% confidence intervals for \gls{BA} model
    parameters estimated with kernel-\gls{ABC}.}
    \label{tab:abchpd}
\end{table}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1.09 
  [0.6 -
   1.16];
$I$,
  1103 
  [662 -
   4455];
$m$,
  3 
  [2 -
   5];
$N$,
  12582 
  [3590-
   14977].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.39 
  [0 -
   0.62];
$I$,
  2360 
  [1379 -
   3811];
$m$,
  2 
  [2 -
   5];
$N$,
  13537 
  [2855 -
   14649].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

\subsubsection*{Effect of parameters on power-law exponent}

\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
to the observed distribution of \gls{gamma} values. The coefficients are
interpretable as multiplicative effects.

\begin{table}
    \centering
    \input{\tablepath/pa-gamma-glm}
    \caption{Estimated \gls{GLM} parameters for relationship between power-law
    exponent \gls{gamma} and \gls{BA} model parameters.}
    \label{tab:glm}
\end{table}

@@

We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
for each dataset are shown in
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
estimates [95\% HPD] 
  1.05 
  [0.04 - 
   1.27]
and
  0.84 
  [0.01 -
   1.02] respectively).
The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (0.04 - 
   1.27).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (0.06 
  [0.01 - 
  0.73]
and
  0.19 
  [0.01 -
   0.8]). 
However, the MAP estimates increased when $m = 1$ was not permitted, although
the HPD intervals remained roughly the same
  (0.78 
  [0.02 - 
  0.94]
and
  0.59 
  [0.07 -
   0.95]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (0.32 for $m \geq 1$;
   0.39 for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([0.04 -
    1.62] for $m \geq 1$ vs.
    0 -
    0.73 for $m \geq 2$).

@@

\begin{figure*}[ht]
  \centering
  \includegraphics{realdata-hpd-bc}
  \caption{
      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
      parameters of the BA network model, fitted to five published HIV datasets
      with kernel-ABC.
  }
  \label{fig:abchpd}
\end{figure*}

>>>>

The \gls{alpha} parameter was the most accurately estimated, with point
estimates having an average deviation of 
    0.14
from the true value, on a grid from 0 to 2. The error was negatively correlated
with the true value of \gls{alpha} 
    (Spearman's $\rho$ = \ensuremath{-0.26},
     $p = 10^{-5}$),
although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
The accuracy was highest for the test value \gls{alpha} = 1.25
    (mean error 0.02)
which exhibited markedly different behaviour than the other values in terms of
the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
particular, there was a very pronounced peak in scores around the true value,
in contrast to most other values where the scores were flat around the true
value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
average absolute error of the point estimates for \gls{I} was 
    310,
on a grid of 500 to 5000, and the errors were not significantly correlated with
the true value of \gls{I}. Kernel score distributions for all test values
exhibited a similar rounded shape (\cref{fig:gridI}).

@@

We used \software{netabc} to estimate the parameters of the \gls{BA} model on
simulated trees where the true parameter values were known. Point estimates for
each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
2. The results for the other values of \gls{m} were similar
(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates of
\gls{alpha} across all simulations was
    0.08 
    [0.05-0.17].
The accuracy of the estimates was not significantly different between values of
$m$ or $I$ (both one-way ANOVA,
    $p$ = 0.05
and 
    0.07),
although the errors when the true value of \gls{alpha} was zero were
significantly greater than the other values 
    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$).
The error in the estimated value of $I$ was
    395 
    [207-683].
Errors were significantly higher for $\alpha \geq 1$
    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$)
and for $I$ = 2000
    ($p = {<}10^{-5}$),
but not for any values of $m$ (one-way ANOVA). The $m$ parameter was estimated
correctly in only
    27 \%
of simulations, little better than random guessing. The true values of $m$ and
$I$ did not significantly affect the error (one-way ANOVA). Finally, the total
number of nodes \gls{N} was consistently over-estimated by about a factor of
two
    (error 5987 
    [2060 - 
     7999]).
No other parameters influenced the accuracy of the $N$ estimates (one-way
ANOVA).

@@

\begin{figure}
  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
  \caption[Approximate marginal posterior distributions of BA model parameters
      obtained by applying \textit{netabc} to a simulated transmission tree
      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
    {
        Approximate marginal posterior distributions of BA model parameters
        obtained by applying \textit{netabc} to a simulated transmission tree
        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
        highest posterior density intervals. $x$-axes indicate regions of
        nonzero prior density.
    }
  \label{fig:abcex}
\end{figure}

\begin{table}
    \centering
    \input{\tablepath/abc-hpd}
    \caption[
        Maximum \textit{a priori} estimates and 95\% highest posterior density
        (HPD) interval boundaries for \gls{BA} model parameters estimated with
        \software{netabc}, averaged over simulated transmission trees.
    ]
    {
        Maximum \textit{a priori} estimates and 95\% highest posterior density
        (HPD) interval boundaries for \gls{BA} model parameters estimated with
        \software{netabc}, averaged over simulated transmission trees.
    }
    \label{tab:abchpd}
\end{table}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1.1 
  [0.6 -
   1.16];
$I$,
  1091 
  [662 -
   4455];
$m$,
  3 
  [1 -
   5];
$N$,
  12566 
  [3733 -
   14945].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.37 
  [0 -
   0.62];
$I$,
  2407 
  [1423 -
   3811];
$m$,
  3 
  [2 -
   5];
$N$,
  13811 
  [2624 -
   14647].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

%\subsubsection*{Effect of parameters on power-law exponent}
%
%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
%to the observed distribution of \gls{gamma} values. The coefficients are
%interpretable as multiplicative effects.
%
%\begin{table}
%    \centering
%    \input{\tablepath/pa-gamma-glm}
%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
%    exponent \gls{gamma} and \gls{BA} model parameters.}
%    \label{tab:glm}
%\end{table}

@@

We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
for each dataset are shown in
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
estimates [95\% \gls{HPD}] 
  1.05 
  [0.04 - 
   1.27]
and
  0.84 
  [0.01 -
   1.02] respectively).
The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (0.04 - 
   1.27).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (0.06 
  [0.01 - 
  0.73]
and
  0.19 
  [0.01 -
   0.8]). 
However, the MAP estimates increased when $m = 1$ was not permitted, although
the HPD intervals remained roughly the same
  (0.78 
  [0.02 - 
  0.94]
and
  0.59 
  [0.07 -
   0.95]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (0.32 for $m \geq 1$;
   0.39 for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([0.04 -
    1.62] for $m \geq 1$ vs.
   [0 -
    0.73] for $m \geq 2$).

@@

\begin{figure*}[ht]
  \centering
  \includegraphics{realdata-hpd-bc}
  \caption[
      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
      parameters of the BA network model, fitted to six HIV datasets
      with \software{netabc}.]
  {
      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
      parameters of the BA network model, fitted to six HIV datasets with
      \software{netabc}. Legend labels indicate risk group and country of
      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
      with men; HET, heterosexual.
  }
  \label{fig:abchpd}
\end{figure*}
